\chapter{More lattices}

After reading the LLL chapter, the reader might have possibly come to the conclusion that Regev collects a single sample - the first vector from LLL - from every run of the algorithm, and hopes that this vector happens to correspond to a non-trivial root. That is, however, not the case. Certain procedures we've described up to now, like the gaussian superposition and the chance of hitting a trivial root rely on probabilistic outcomes, and we also need to account for quantum noise. Regev therefore attempts to generate as many vectors as possible from a single run, to increase his probability of finding non-trivial roots.

Now, suppose we have our $d+k$ d-dimensional samples of $\mathit \Lambda^*$ $\mathbf{v_1,v_2}\dots \mathbf{v_{d+k}}$ (by normalized, we mean that we have divided the original samples by $D$ and now every entry is in $[0,1)$. Assuming our samples aren't exactly accurate, there has to be some additive noise $\delta_i$ for each $\mathbf{v_i}$. Let $\delta$ be some upper bound on the magnitude of the $\delta_i$, and define the $2d+k$ dimensional lattice $\mathit \Lambda'$ spanned by the columns of the following matrix (we need $k \geq 4$ from Pomerance)
\begin{equation}
M = \left[
\begin{array}{c|c}
  I_{d\times d} & 0 \\
  \hline
  \begin{matrix} \delta^{-1}\mathbf{v_1} \\  \vdots \\ \delta^{-1}\mathbf{v_{d+k}} \end{matrix} & \delta^{-1} \cdot I_{k\times k}
\end{array}
\right]    
\end{equation}


Regev proves using the Cauchy-Schwarz inequality the following lemma (combination of Lemma 4.4 and Corollary 4.5 in \cite{Regev2025})\footnote{We omit the proofs leading up to this result due to them being lengthy and hard to explain in simple terms.} \\
\textbf{Lemma 10.1}: For any $\mathbf{u} \in \Lambda$, there exists a vector $\mathbf{u}' \in \Lambda'$ whose first d coordinates are equal to \textbf{u} with norm $\lVert \mathbf{u'} \rVert \leq \lVert \mathbf{u} \rVert \cdot \sqrt{1+d+k}$. Moreover, for $k = 4 $ it holds that the first $d$ coordinates of every nonzero vector $\mathbf{u'} \in \Lambda'$ with norm $$\lVert \mathbf{u'} \rVert < \frac{\delta^{-1}}{6 \cdot (4\cdot det(\Lambda))^{1/d+4}}$$ form a vector in $\Lambda$ with probability at least 1/4.

From this correspondence, we can see how from our noisy samples of $ \mathit \Lambda^*$ we will produce a clean, short base for $\mathit \Lambda$ sufficient to give us a non-trivial factor. Suppose we run LLL on the $\mathit \Lambda'$ basis formed by \textbf{Eq.10.1}, returning some vectors $\mathbf{b_1, b_2, \dots, b_{d+k}}$ and then Gram-Schmidt orthogonalization on the result, giving us some orthogonal vectors $\mathbf{b^*_1, b^*_2, \dots, b^*_{d+k}}$. Let $T$ be some norm bound, and let $h \leq k$ be the largest number such that 
\begin{equation}
    \lVert\mathbf{b^*_h} \rVert <  2^{k/2} T
\end{equation}
A corollary of the Lovasz condition can help tighten this bound to\footnote{It should be obvious that all $b^*_i$ with $i\leq h$ also obey this bound.}. \begin{equation}
    \lVert\mathbf{b^*_h} \rVert <  \sqrt{k}2^{k/2}T
\end{equation}
Then, according to Claim 5.1 in Regev's paper, any vector in the sublattice $\mathit \Lambda'_T \subseteq \mathit \Lambda'$ formed by vectors of norm $\leq T$ can be generated by $\mathbf{b_1}, \mathbf{b_2}, \dots ,\mathbf{b_h}$. 
Due to Lemma 10.1, the first d coordinates of the $\mathbf{b_i}$ correspond to vectors in $\mathit \Lambda$. This gives us not just one, but multiple short vectors in $\mathit \Lambda$ that could also be in $\mathit \Lambda / \mathit \Lambda_0$. In fact, at least one of those vectors is guaranteed to not be in $\mathit \Lambda_0$, otherwise $\mathit \Lambda_T$  would be equal to $\mathit \Lambda_0$ itself (since $\mathit \Lambda_0$ is a group with closure). Therefore, taking $k \geq 4$ (to overshoot the >1/4 probability that each one of them is in $\mathit \Lambda_0$), we can reliably end up with a vector corresponding to a non-trivial factor.












