\chapter{Multiplication of small numbers}


Regev further presents us with a potential efficient way we could compute the product $$X =\prod_{i=1}^d b_i^{e_i} \pmod N$$ To do this, he combines two ideas that stem from fast multiplication algorithms.

$\textbf{Lemma 7.1.}$ There exists a way to multiply $d$ small numbers using $O(d \cdot (logd)^k)$ multiplications for some $k$ in a way that is more efficient than sequential multiplication. 

Say we have e.g. $a_1\dots a_8$ in ascending order of value, and we want the product $a_1a_2a_3a_4a_5a_6a_7a_8$. The naive way of computing it would be to multiply $a_1a_2$, then product $a_1a_2$ with $a_3$ and so on. This, however, is unoptimal, because our sequential chain will soon reach n bits, and from then on we will be multiplying n-bit numbers (registers need to match for multiplication, so we have to use the larger register). Instead, it is much more efficient to do $$((a_1a_8)(a_2a_7))((a_3a_6)(a_4a_5))$$ in a binary tree fashion, since you are always multiplying small numbers with small numbers, and we only hit n-bit numbers near the root of the tree. 

$\textbf{Lemma 7.2}$.Modular exponentiation can be performed in $O(logk)$. 

Suppose we need to compute some large power $k$ of a small number $x$ modulo N where $N > 1$, so $x^k \pmod N$. Let for example $k = 13$. To compute $x^{13}$, we work as follows.

$\mathbf{2.1}$ Define two classical registers. One will hold the base, and another will hold the running total. Initialize the base register to $x \pmod n$, and the running total to 1.\\
$\mathbf{2.2}$ Find the binary representation of 13, which is $1101_2$. We will have as many steps as there are numbers in the binary representation, in this case 4.\\
$\mathbf{2.3}$ Examining the digits \textbf{right to left}, do the following:\\
$\mathbf{2.3.1}$ Look at the digit.If the digit is 1, we multiply the running total by the base register and store the result in the running total. If the digit is 0, we do nothing.\\ 
$\mathbf{2.3.2}$ If this is the last digit,terminate the algorithm.\\
$\mathbf{2.3.3}$ Square the base register.\\
$\mathbf{2.3.4}$ Repeat from 2.3.1 until there are no more digits left.\\

Below is pseudocode of the algorithm. 

\begin{algorithm}[H]
\caption{Modular Exponentiation}
\textbf{Input:} Integers $x,k,m$ \\
\textbf{Output:} $x^k \pmod m$
\begin{algorithmic}[1]
\State $b \gets x \pmod m$
\State $rt \gets 1$
\While{$k > 0$}
    \If {$k \equiv 1 \pmod 2$}
        $rt \gets rt \cdot b \pmod m$
    \EndIf
    \State $k \gets k /2$ (integer division)
    \If {$k = 0$}
        \State \textbf{break;}
    \EndIf
    \State $b \gets (b \cdot b) \pmod m$
\EndWhile
\State \Return $rt$
\end{algorithmic}
\end{algorithm}

This speeds up exponentiation modulo m from requiring $k-1$ multiplications in the naive case, to $O(logk)$. Now, let us combine the two ideas into an algorithm that efficiently computes $X$.\\ 

We will describe the way the algorithm would compute a product like $a_1^{13} a_2^9 a_3^3 a_4^6 \pmod m$.The binary representations of the exponents are as follows $13 = 1101_2$ $9 = 1001_2$ $3 = 0011_2$ $6 = 0110_2$. This time,we will only use one register $rt$ for the result, which we we will initialize to 1.\\

\textbf{Step 1.}Examine the first bit. $13$ and $9$ have 1, so $rt \gets a_1a_2 \cdot rt \pmod m$.\\
\textbf{Step 2.}Square the result,so $rt \gets rt \cdot rt \pmod m$.\\
Current state of result: $rt = a_1^2a_2^2$.\\
\textbf{Step 3.}Examine the second bit. $13$ and $6$ have 1,so $rt \gets a_1a_4 \cdot rt \pmod m$.\\
Current state of result:$rt = a_1^3a_2^2a_4$.\\
\textbf{Step 4.}Square the result.\\
Current state of result: $rt = a_1^6a_2^4a_4^2$.\\
\textbf{Step 5.}Examine the third bit. $3$ and $6$ have 1,so $rt \gets a_3a_4 \cdot rt \pmod m$.\\
Current state of result:$rt = a_1^6a_2^4a_3a_4^3$.\\
\textbf{Step 6.}Square the result.\\
Current state of result: $rt = a_1^{12}a_2^8a_3^2a_4^6$.\\
\textbf{Step 7.}Examine the fourth bit. $13,9$ and $3$ have 1,so $rt \gets a_1a_2a_3 \cdot rt \pmod m$.\\
Current state of result: $rt = a_1^{13}a_2^9a_3^3a_4^6$.\\
\textbf{Step 8.}No more bits in the binary representation, so the algorithm terminates.\\

The intermediate multiplication step can be done efficiently in a quantum setting by using the binary tree that contains all the possible bases as leaves, and computing the final multiplication factor as outlined in Lemma 1, using controlled multiplications. Since a tree with $d$ leaves has $d - 1$ interior nodes,letting $d \approx \sqrt{n}$ - this is the value we will be using in the algorithm- we would need $(d + d-1) \cdot n = O(2n^{3/2})$ qubits, plus $O(n)$ ancillas, for out of place multiplication (meaning we have three registers: the two multiplicands and the result) - this is the implementation we will be using in the Qiskit code for the algorithm.

We can optimize that by seeing that not every register needs to have size n. We can compute exactly how many qubits each of our product registers will need by simply summing the length of the respective multiplicands, and allocate them accordingly. In Regev's algorithm, the bases are the squares of the first d primes, and so, using the Prime Number Theorem, an upper bound for the total size of the first layer is $$\# qubits < 2\cdot 1.01624\cdot p_d \cdot log_2{e}$$ which can be rewritten as
\begin{equation}
    \# qubits < 2.93p_d
\end{equation}

where $p_d$ is the d-th prime number.

The gate complexity here is approximately $O(n^{1.5} poly(logn))$. To see this, note that in Chapter 6 we assumed that $e_i <= e^{O(d)} \approx 2^O(d)$, and therefore the largest exponent in our expression has $O(d)$ bits. The steps of the algorithm depend on the bit length of the largest exponent, so we have $O(d) = O(\sqrt{n})$ steps. Multiplication can be done in $O(n) poly(logn)$. Combining these, we get the desired result.\\

It should be noted here that most optimizations of Regev's algorithm that have come out to this day are concerned with optimizing this specific operation in terms of gates and/ or qubits. We describe the two most important ones in chapters 14 and 15. 




