\chapter{Regev optimizations: Fibonacci exponentiation}

The first optimization we are going to talk about has to do with transforming the base in which we represent the numbers in a base that is perhaps more suited when dealing with exponents. To do this, we will entice the amateur mathematics enjoyer by bringing the Fibonacci numbers and sequence into the picture.
To explore this concept, first we will need to talk about Zeckendorf's theorem.The statement is the following:\\

$\textbf{Lemma 15.1 (Zeckendorf)}$\\
Every positive integer $k$ can be uniquely represented as a sum of non-consecutive Fibonacci numbers.In other word, for every $k \in \mathbb{Z}$:
\begin{equation}
    k = \sum_j c_jFi, \: c_j \in \set{0,1}
\end{equation}

No two consecutive $c_i$ can be 1, because consecutive Fibonacci numbers can be replaced with the next Fibonacci number according to the recursive formula $F_n = F_{n-1} + {F_n-2}$. This means that representing exponents as bits using a Fibonacci base is perfectly valid. We will now describe a way of computing $a^{k}$ using a Fibonacci base.First, we need to compute the Fibonacci representation of k. To do that in a quantum circuit, we use a greedy algorithm.

$\textbf{Step 1.}$ Compute all Fibonacci numbers up to $k$ using the recursive formula, starting from $F_1 = 1,F_2 = 2$. Initialize an array F of 0s with size m-1, where $F_m$ is the largest Fibonacci number smaller than $k$ and let $j \gets m - 1$.\\
$\textbf{Step 2.}$ $F[j] = 1$\\
$\textbf{Step 3.}$ $k \gets F_m - k$\\
$\textbf{Step 4.}$ Compute the largest Fibonacci number smaller than k using binary search on the computed Fibonacci series\\
$\textbf{Step 5.}$Repeat from Step 2 until $k = 0$.\\


F contains the binary representation of k. Now, computing $a^k$ is very easy, because we can use the same logic as modular exponentiation.\\

Assume the greedy algorithm in the previous step leaves us with a Fibonacci representation $k_{\phi}$ consisting of all the $c_j$, and $m$ Fibonacci numbers computed, starting from $F_2 = 1$. Let us present the algorithm for Fibonacci exponentiation, as described by Kaliski \cite{Kaliski2017}\\

\begin{algorithm}
\caption{Fibonacci exponentiation}
\textbf{Input:} $a, k, N$, the sequence $c_1,c_2 \dots c_m$\\
\textbf{Output:} $a^k \pmod N$
\begin{algorithmic}[1]
\State $(c,d,e) \gets (1,a,1)$
\For{$j = 1$ to $m$}
    \State $(c,d) \gets (d,cd \pmod N)$
    \If{$c_j == 1$}
        \State $e \gets de \pmod N$
    \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

This algorithm, however, poses a problem. First, note that gates in a quantum circuit must  necessarily be reversible, else there is no way of undoing a computation to in order to clear the quantum register. Step 3 on a quantum level means performing the computation $$(a, b) \mapsto(a, ab\mod N)$$. The inverse gate is $$(a, b) \mapsto(a, a^{-1}b \mod N)$$ Setting $b = 1$, this is equivalent to computing an inverse modulo N, meaning Step 3 requires a computation that is at least as hard as that. Note that in Regev's algorithm we severely limited how many of those operations we had to perform for the same reason - it's not easy \footnote{see section 1.3 in Ragavan and Vaikuntanathan's paper \cite{RagavanVaikuntanathan2023} for more information}.\\



Consider the original Regev product $\prod_{i=1}^d b_i^{e_i}$ (the algorithm is also compatible with the $D/2$ shift). Suppose we have a fibonacci representation of $e_i$ 
\begin{equation}
    \prod_{j=1}^m c_{i,j} F_j
\end{equation}
where $c_{i,j}$ denotes each of the earlier $c_j$ but now for every base. Then we can plug \textbf{Eq 15.2} into the Regev product and write the resulting expression - which is \textbf{equivalent} to the Regev product, as follows: $$\prod_{j=1}^m(\prod_{i=1}^d b_i^{c_{i,j}})^ {F_j}$$
Denote the inner product as $C_j$. This leaves us with
\begin{equation}
    \prod_{j=1}^m C_j^ {F_j}
\end{equation}
Then, the inverse of $C_j$ modulo N is 
\begin{equation}
    C_j^{-1} = \prod_{i=1}^d b_i^{-c_{i,j}} = \prod_{i=1}^d b_i^{-1} \cdot \prod_{i=1}^d b_i^{1 -c_{i,j}}
\end{equation}
The first term is a constant that can be precomputed classically, since we decide what the $b_i$ will be before the algorithm starts, using the extended Euclidean algorithm. Since $c_{i,j} \in \set{0,1}$, the second term is also easy to compute by applying Algorithm 3 with all $c_i$ bit flipped. Given a precomputation of the modular inverses of the bases $b_i$ modulo N, in Lemmas 5.1 and 5.2 of \cite{RagavanVaikuntanathan2023}, we are provided with the proof for the existence of a circuit that, given $a, b, a^{-1},b^{-1}$ in quantum registers, produces quantum registers that include $ab$ alongside $(ab)^{-1}$. This means we can drag $C_j$ and $C_j^{-1}$ along for the entirety of the circuit.

Let us, finally, describe Algorithm 5.2 from Ragavan and Vaikuntanathan\cite{RagavanVaikuntanathan2023} in terms of what we wrote earlier. It is important to note that this is applied to the shifted exponent $e_i = c_i + D/2$.

\begin{algorithm}[H]
\caption{Fibonacci multi-exponentiation (top-down)}
\textbf{Input:} Initial quantum state $\ket{e}$, precomputed $B = \prod_{i=1}^d b_i^{-1}$\\
\textbf{Output:} State consisting of a superposition $\prod_{i=1}^d b_i^{e_i} \pmod N$
\begin{algorithmic}[1]
\State Compute $c_{i,j}$ according to our initial algorithm.
\State $R_1 \gets 1$, $R_2 \gets 1$
\For {$j = m$ downto 1}
    \State $R_1 \gets R_1R_2$
    \State Compute $\ket{\prod_{i=1}^d b_i^{1 -c_{i,j}}}$
    \State Use $B$ to convert the previous state into $\ket{C_j^{-1} \pmod N}$
    \State Compute the state $\ket{C_j}$. 
    \State $R_1 \gets R_1C_j$
    \State Uncompute $\ket{C_j}$ using $\ket{C_j^{-1}}$
    \State Swap $R_1$ and $R_2$ (the registers that contain the superpositions created for both)
\EndFor
\State $R_2$ is now entangled with $\ket{e}$, and holds the superposition of the possible products.
\State \Return $R_2$
\end{algorithmic}
\end{algorithm}

We can think of $C_j$ as a multiplication factor that is applied in each individual step, just as Regev would apply, for example, $b_1b_3b_4$. It is very important that we go top-down, because it means that when inverting the multiplication factor, the values we remove aren't required for the later registers - they are used up and can be safely "destroyed" by inversion without affecting the entangled superposition. The operation performed in each step is essentially $R_1 \gets R_1 R_2 C_j$, so we accumulate the existing product by the next Fibonacci number and the multiplication factor, and then we swap the registers so that $R_1$ always has the previous value at the end of the loop.

We avoided thorough analysis of the qubit requirements here because, although they constitute a significant part of why everything is done the way it is, explaining them thoroughly would take incredibly long. All that needs to be said is that, given that a black box multiplication circuit that performs
$$\ket{a} \ket{b} \ket{t} \ket{0^S} \to \ket{a} \ket{b} \ket{t+ ab} \ket{0^S}$$ requires $S = log_2N + A$ qubits ($A$ refers to the ancillas) and $G$ gates, the above algorithm can be executed with $S + (A/log\phi + 6 + O(1))n$ qubits and $O(\sqrt{n}G + n^{3/2})$ gates, which for the classical Van Der Hoeven algorithm for multiplication\cite{HarveyVanDerHoeven2021} is equivalent to $O(nlogn)$ qubits and $O(n^{3/2} logn)$ gates, achieving a significant improvement over Regev's $O(n^{3/2})$ qubits.






